### 2024-03-26 doc/sections/mav-semantics.tex
[ tex ] semantics semantics semantics

@@

\begin{definition}
  An \emph{MAV-algebra} is a structure
  $(A, \leq, \otimes, ;, I, \lnot)$ with the following properties:
  \begin{enumerate}
    \item $(A, \leq, \otimes, I, \lnot)$ is $*$-autonomous and satisfies \emph{mix}.
    \item $(A, \leq, ;, I)$ is a pomonoid.
    \item $(\otimes, I)$ and $(;, I)$ are in a duoidal relationship.
    \item $(A, \leq)$ has binary meets, which we write as $x \with y$.
  \end{enumerate}
\end{definition}

@@

To prove completeness of the cut-free fragment of SMAV, we will
construct a particular MAV-algebra from the syntax and cut-free
proofs. Since cut-free proofs do not a priori have all the necessary
structure for an MAV-algebra, we develop a procedure to construct
MAV-algebra from the lighter requirements of an MAV-frame, defined in
this section. MAV-frames have an intuitive reading as a kind of
process algebra, so we use process algebra-like notation. In the
following sections, we show how every MAV-frame induces an
MAV-algebra, and in Section XXX we will show that the MAV-algebra
constructed from the cut-free proof MAV-frame allows us to prove that
all proofs in SMAV have cut-free normal forms.

\begin{definition}
  An \emph{MAV-frame} is a structure
  $(F, \leq, I, \parallel, \lhd, +)$ where $(F, \leq)$ is a partial
  order, $(F, \leq, I, \parallel)$ is a commutative pomonoid,
  $(F, \leq, I, \lhd)$ is a pomonoid, $+$ is a binary monotone
  function on $(F, \leq)$, and these structures satisfy the following
  inequalities:
  \begin{enumerate}
    \item $(w \lhd x) \parallel (y \lhd z) \leq (w \parallel y) \lhd (x \parallel z)$
    \item $(x + y) \parallel z \leq (x \parallel z) + (y \parallel z)$
    \item $(w \lhd x) + (y \lhd z) \leq (w + y) \lhd (x + z)$
    \item $I + I \leq I$
  \end{enumerate}
\end{definition}

@@

\begin{proposition}
  \bob{The Cut-free system forms an MAV-frame; fwd ref to next section}
\end{proposition}

@@

Formal $+$-contexts are the semantic analogue of Horne's \emph{Killing
  contexts}.

\begin{definition}
  Let $(A, \leq)$ be a partial order with a monotone binary operation
  $+ : A \times A \to A$.  Let $F \in \LowerSet{A}$. A \emph{formal
    $+$-context} over $F$ is a binary tree built from constructors
  $\mathsf{leaf} : F \to \mathrm{ctxt}(F)$ and
  $\mathsf{node} : \mathrm{ctxt}(F) \times \mathrm{ctxt}(F) \to
    \mathrm{ctxt}(F)$. We define the \emph{sum} of a context as
  $\mathit{sum}(\mathsf{leaf}~x) = x$ and
  $\mathit{sum}(\mathsf{node}(c,d)) = \mathit{sum}(c) +
    \mathit{sum}(d)$.
\end{definition}

@@

\begin{proposition}
  $(\ClosedLowerSet{A}, \subseteq)$ has all meets and joins. In the
  binary case, meets are defined by intersection and joins are defined
  by $x \lor y = \alpha (U(x) \cup U(y))$.
\end{proposition}

@@

\begin{proposition}\label{prop:closed-monoid-distrib}
  \bob{Monoid construction 1, for monoids that distribute over
    $+$. This also has a residual.}
\end{proposition}

@@

\begin{remark}
  We have used the same decoration $\ClosedDay{\bullet}$ and
  $\ClosedDay{\lhd}$ for two separate constructions of pomonoids on
  $+$-closed lower sets. We will be careful to distinguish which we
  mean: a symmetric operator like $\bullet$ will distribute over $+$
  and so $\ClosedDay{\bullet}$ will be constructed by Proposition
  \ref{prop:closed-monoid-distrib}; and a non-symmetric operator like
  $\lhd$ will be entropic with respect to $+$ and so
  $\ClosedDay{\lhd}$ will be constructed by Proposition
  \ref{prop:closed-monoid-duoidal}.
\end{remark}

\begin{remark}
  If we have two $(A, \leq)$ pomonoids with the same unit, then the
  two constructions will yield the same closed lower set.
  \bob{elaborate a bit...}
\end{remark}

@@

To construct suitable MAV-algebras, we use the poset version of the
Chu construction \cite{barr}. The Chu construction builds
$*$-autonomous categories from symmetric monoidal closed categories
with pullbacks. In the preorder case, the requirement for preorders
simplifies to binary meets. Therefore, for this section, we let
$(A, \leq, \bullet, i, \rightblackspoon)$ be a symmetric monoidal
closed preorder with all binary meets.

@@

\begin{proposition}
  $\Chu(A, k)$ is a $*$-autonomous partial order with pomonoid and
  negation structure defined as:
  \begin{mathpar}
    (a^+, a^-) \otimes (b^+, b^-) = (a^+ \bullet b^+, (b^+ \rightblackspoon a^-) \land (a^+ \rightblackspoon b^-))

    I = (i, k)

    (a^+,a^-)^\perp = (a^-, a^+)
  \end{mathpar}
\end{proposition}

\begin{remark}
  If we choose $k = i$, then $\Chu(A, i)$ is a $*$-autonomous partial
  order that satisfies \emph{mix}.
\end{remark}

\begin{proposition}
  If $A$ has binary joins, then $\Chu(A, k)$ has binary meets, given
  by $(a^+,a^-) \with (b^+,b^-) = (a^+ \land b^+, a^- \lor b^-)$.
\end{proposition}

\paragraph{Self-dual duoidal structure}
These two theorems give us enough structure to interpret the
multiplicative and additive constructs of MAV. We need to also
interpret the self-dual $P; Q$ connective. Let us assume that we have
an operation $-\lhd- : A \times A \to A$ that we want to induce
self-dual operator on $\Chu(A, k)$. By the way the Chu construction is
defined, we can make an operator that is self-dual by definition:
\begin{displaymath}
  (a^+, a^-) ; (b^+, b^-) = (a^+ \lhd b^+, a^- \lhd b^-)
\end{displaymath}
for this to work, we need to assume duoidality.

And that the dualising object $k$ is a
satisfies $k \lhd k \leq k$. If $\lhd$ has a unit $j$, then $(j, j)$ is a unit if

\begin{proposition}\label{prop:chu-self-dual}
  Let $(\lhd, j)$ be a pomonoid on $(A, \leq)$ such that
  $k \lhd k \leq k$ and $j \leq k$, then
  $x ; y = (x^+ \lhd y^+, x^- \lhd y^-)$ and $J = (j, j)$ form a
  self-dual pomonoid on $\Chu(A, k)$.
\end{proposition}

@@

\begin{lemma}
  \bob{entropy on residuals}
\end{lemma}

@@

\begin{theorem}
  If $(F, \leq, \parallel, \lhd, I, +)$ is an MAV-frame, then
  $\Chu(\ClosedLowerSet{F}, I)$ has the structure of an MAV-algebra.
\end{theorem}

>>>>

\begin{definition}
  \bob{residuation}
\end{definition}

@@

\begin{definition}
  An \emph{MAV-algebra} is a structure
  $(A, \leq, \otimes, ;, I, \lnot)$ with the following properties:
  \begin{enumerate}
  \item $(A, \leq, \otimes, I, \lnot)$ is $*$-autonomous and satisfies \emph{mix}.
  \item $(A, \leq, ;, I)$ is a pomonoid.
  \item $;$ is self dual: $\lnot (x ; y) = (\lnot x); (\lnot y)$.
  \item $(\otimes, I)$ is duoidal over $(;, I)$.
  \item $(A, \leq)$ has binary meets, which we write as $x \with y$.
  \end{enumerate}
\end{definition}

@@

To prove completeness of the cut-free fragment of SMAV, we will
construct a particular MAV-algebra from the formulas and cut-free
proofs. Since cut-free proofs do not a priori have all the necessary
structure for an MAV-algebra, we develop a procedure to construct
MAV-algebra from the lighter requirements of an MAV-frame, defined in
this section. MAV-frames have an intuitive reading as a kind of
process algebra, so we use process algebra-like notation. In the
following sections, we show how every MAV-frame induces an
MAV-algebra, and in Section XXX we will show that the MAV-algebra
constructed from the cut-free proof MAV-frame allows us to prove that
all proofs in SMAV have cut-free normal forms.

\begin{definition}
  An \emph{MAV-frame} is a structure
  $(F, \leq, I, \parallel, \lhd, +)$ where $(F, \leq)$ is a partial
  order, $(F, \leq, I, \parallel)$ is a commutative pomonoid,
  $(F, \leq, I, \lhd)$ is a pomonoid, $+$ is a binary monotone
  function on $(F, \leq)$, and these data satisfy the following
  inequalities:
  \begin{enumerate}
    \item $(w \lhd x) \parallel (y \lhd z) \leq (w \parallel y) \lhd (x \parallel z)$
    \item $(x + y) \parallel z \leq (x \parallel z) + (y \parallel z)$
    \item $(w \lhd x) + (y \lhd z) \leq (w + y) \lhd (x + z)$
    \item $I + I \leq I$
  \end{enumerate}
\end{definition}

@@

\begin{proposition}
  The \emph{cut-free MAV-frame} $\textsc{CFMav}$ is the partial order
  arising as the quotient of the preorder formed from the formulas of
  MAV and $P \leq Q$ if $P \longrightarrow^* Q$. Define
  $P \parallel Q = P \vParr Q$, $P \lhd Q = P \vSeq Q$,
  $P + Q = P \vWith Q$ and $I = \vUnit$. The required (in)equalities
  follow directly from the definition of $\longrightarrow^*$.
\end{proposition}

\begin{remark}
  The construction of the MAV-frame \textsc{CFMav} does not use the
  $\otimes$ and $\oplus$ structure of MAV directly. We will see in
  Section \ref{sec:mav-cut-elimination} that this structure is
  recovered by duality from the other connectives by the constructions
  in the rest of this section. \bob{Which one is ``positive'' again?}
\end{remark}

@@

$+$-contexts are the semantic analogue of Horne's \emph{Killing
  contexts}.

\begin{definition}
  Let $(A, \leq)$ be a partial order with a monotone binary operation
  $+ : A \times A \to A$.  Let $F \in \LowerSet{A}$. A
  \emph{$+$-context} over $F$ is a binary tree built from constructors
  $\mathsf{leaf} : F \to \mathrm{ctxt}(F)$ and
  $\mathsf{node} : \mathrm{ctxt}(F) \times \mathrm{ctxt}(F) \to
    \mathrm{ctxt}(F)$. We define the \emph{sum} of a context as
  $\mathit{sum}(\mathsf{leaf}~x) = x$ and
  $\mathit{sum}(\mathsf{node}(c,d)) = \mathit{sum}(c) +
    \mathit{sum}(d)$.
\end{definition}

@@

\begin{proposition}
  $(\ClosedLowerSet{A}, \subseteq)$ has all meets and joins. In the
  binary case, meets are defined by intersection and joins are defined
  by $F \lor G = \alpha (U F \cup U G)$.
\end{proposition}

@@

\begin{proposition}\label{prop:closed-monoid-distrib}
  For a commutative pomonoid $(\bullet, i)$ on $(A, \leq)$ that
  distributes over $+$
  ($(x \bullet y) + z = (x + z) \bullet (y + z)$), then
  $F \ClosedDay{\bullet} G = \alpha(U F \Day{\bullet} U G)$ and
  $\ClosedDay{j} = \alpha(\Day{j})$ define a residuated commutative
  pomonoid on $\ClosedLowerSet{A}$.
\end{proposition}

@@

\begin{remark}
  We have used the same decoration $\ClosedDay{\bullet}$ and
  $\ClosedDay{\lhd}$ for two separate constructions of pomonoids on
  $+$-closed lower sets. We will be careful to distinguish which we
  mean: in our present application, a symmetric operator like
  $\bullet$ will distribute over $+$ and so $\ClosedDay{\bullet}$ will
  be constructed by Proposition \ref{prop:closed-monoid-distrib}; and
  a non-symmetric operator like $\lhd$ will be entropic with respect
  to $+$ and so $\ClosedDay{\lhd}$ will be constructed by Proposition
  \ref{prop:closed-monoid-duoidal}.
\end{remark}

\begin{remark}
  If we have two pomonoids on $(A, \leq)$ that share a unit, then the
  two constructions of units in Propositions
  \ref{prop:closed-monoid-distrib} and
  \ref{prop:closed-monoid-duoidal} will yield the same element of
  $\ClosedLowerSet{A}$.
\end{remark}

@@

To construct suitable MAV-algebras, we use the poset version of the
Chu construction \cite{barr}. The Chu construction builds
$*$-autonomous categories from symmetric monoidal closed categories
with pullbacks. In the partial order case, the requirement for
pullbacks simplifies to binary meets. Therefore, for this section, we
let $(A, \leq, \bullet, i, \rightblackspoon)$ be a partial order with
a residuated pomonoid structure and all binary meets.

@@

\begin{proposition}
  $(\Chu(A, k), \sqsubseteq)$ has a $*$-autonomous structure defined
  as:
  \begin{displaymath}
    (a^+, a^-) \otimes (b^+, b^-) = (a^+ \bullet b^+, (b^+ \rightblackspoon a^-) \land (a^+ \rightblackspoon b^-))
    \qquad
    I = (i, k)
    \qquad
    \lnot(a^+,a^-) = (a^-, a^+)
  \end{displaymath}
\end{proposition}

\begin{remark}
  If we choose $k = i$, then $(\Chu(A, i), \sqsubseteq)$ has
  $*$-autonomous structure that satisfies \emph{mix}.
\end{remark}

\begin{proposition}\label{prop:chu-meets}
  If $A$ has binary joins, then $(\Chu(A, k), \sqsubseteq) x$ has
  binary meets, given by
  $(a^+,a^-) \with (b^+,b^-) = (a^+ \land b^+, a^- \lor b^-)$.
\end{proposition}

\begin{remark}
  Since $(\Chu(A, k), \sqsubseteq)$ is a $*$-autonomous partial order,
  then Proposition \ref{prop:chu-meets} also means that $\Chu(A, k)$
  has all binary joins, with
  $(a^+, a^-) \oplus (b^+, b^-) = (a^+ \lor b^+, a^- \land b^-)$.
\end{remark}

We now turn to the self-dual duoidal structure required to interpret
the $\vSeq$ connective. First we transfer pomonoids from $(A, \leq)$
to self-dual pomonoids on $(\Chu(A, k), \sqsubseteq)$ provided they
interact well with $k$:
\begin{proposition}\label{prop:chu-self-dual}
  Let $(\lhd, j)$ be a pomonoid on $(A, \leq)$ such that
  $k \lhd k \leq k$ and $j \leq k$, then
  $x ; y = (x^+ \lhd y^+, x^- \lhd y^-)$ and $J = (j, j)$ form a
  self-dual pomonoid on $\Chu(A, k)$.
\end{proposition}

@@

Finally, we need to show that if $(\bullet, j)$ is duoidal over
$(\lhd, j)$, then their Chu counterparts are in the same
relationship. Due to the use of residuals in the definition of
$\otimes$, we need the following fact about duoidal residuated
pomonoids:

\begin{lemma}\label{lem:duoidal-residual}
  If $(\bullet, j)$ is duoidal over $(\lhd, j)$ in a partial order
  $(A, \leq)$ and $(\bullet, j)$ has a residual $\rightblackspoon$,
  then
  $(w \rightblackspoon x) \lhd (y \rightblackspoon z) \leq (w \lhd y)
  \rightblackspoon (x \lhd z)$.
\end{lemma}

\begin{remark}
  Lemma \ref{lem:duoidal-residual} is in some sense the
  ``intuitionistic'' version of the duoidal relationship for $\parr$
  arising as the dual of that for $\otimes$ in a $*$-autonomous
  partial order.
\end{remark}

@@

\begin{theorem}
  If $(F, \leq, \parallel, \lhd, I, +)$ is an MAV-frame, then
  $(\Chu(\ClosedLowerSet{F}, I), \sqsubseteq)$ has the structure of an
  MAV-algebra.
\end{theorem}
